Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-10-27 paper/main.tex
concretizing record transform

The topological structure of the data can be expressed programmatically by constructing data types that encapsulate the class of topological structure-.i.e point, line, plane, network, etc. We propose that these data types can be formally specified as objects of a category. 
\begin{definition} 
   An \textbf{category} $\mathcal{C}$ consists of the following \textit{data}: 
\begin{enumerate}
  \item a collection of \textit{objects} $X \textbf{ob}(\mathcal{C})$
  \item for every pair of objects $X, Y \in \textbf{ob}(\mathcal{C})$, a set of \textit{morphisms} $X \xrightarrow{f} Y \in Hom_{\mathcal{C}}(X, Y)$
  \item for every object $X$, a distinct \textit{identity morphism} $X \xrightarrow {id_x} X$ in $Hom_{\mathcal{C}}(X, X)$
  \item a \textit{composition function} $f \in Hom_{\mathcal{C}}(X, Y) \times  g \in Hom_{\mathcal{C}}(Y, Z) \rightarrow g \circ f \in Hom_{\mathcal{C}}(X, Z)$
\end{enumerate}
such that 
\begin{enumerate}
  \item \textit{unitality:} for every morphism $ X \xrightarrow{f} Y$, $f \circ id_x = f = id_y \circ f$
  \item \textit{associativity:} if any three morphisms $f, g, h$ are composable, 
    \begin{equation*}
      \begin{tikzcd}
        X \arrow[r, "f"] \arrow[rrr, "h\circ(g\circ f) = (h\circ g)\circ f"', bend right, dashed] & Y  \arrow[r, "g"] & Z \arrow[r, "h"] & W
        \end{tikzcd}
  \end{equation*}
  then they are associative such that $h\circ(g\circ f) = (h \circ g) \circ f$  \cite{lawvere2009conceptual,riehlCategoryTheoryContext,maclaneCategoriesWorkingMathematician2013,fongInvitationAppliedCategory2019}. 
  \end{enumerate}
\end{definition}
We encapsulate the topological structure of the data as a category $\mathcal{\dbasec}$. The standard construction of a category from a topological space is that it has open set objects $\opensetc$ and inclusion morphisms $\opensetc_i \xrightarrow{\iota} \opensetc_j$ such that $\opensetc_i \subseteq \opensetc_j$\cite{riehlCategoryTheoryContext}. The composability property expresses that inclusion is transitive, while associativity expresses that the inclusion functions can be curried in various equivalent groupings. By formally specifying the properties of the topological structure datatypes as $\mathcal{\dbase}$, we can express that these are the properties that are required as part of the implementation of the data type objects.

@@

\paragraph{\textbf{Topological structure}}
\noindent
The base space transformation is a pointwise continuous map from one open set to another open set in the same base space
\begin{equation}
  \label{eq:atct:morphism:base}
 \dfunchc: \dbasepointc^{\prime}\mapsto \dbasepointc
 \end{equation}
such that $\openset, \openset^{\prime} \subseteq \dbase$. This means $\openset$ and $\openset^{\prime}$ are of the same topology type. To correctly align the sections with the remapped base space, there is a a corresponding section pullback function
\begin{equation}
  \label{eq:atct:morphism:basepull}
  \dfuncpullc \dsectionc \restriction_{\opensetc^{\prime}}: \dsectionc\restriction_{\opensetc^{\prime}} \mapsto \dsectionc \restriction_{\opensetc^{\prime} \circ \dfunchc} 
\end{equation}
such that $\dsection|_{\openset} = \dfuncpull\dsection|_{\opensetc^{\prime}}$ because $\dsection|_{\openset} = \dsection|_{\dfunch(\opensetc^{\prime})}$. This means that the base space transformation \dfunch\ does not change the data values at a given point
\begin{equation}
  \label{eq:atct:morphism:verify_base}
  \dsection(\dbase) = \dfuncpull\dsection(\dbasepoint^{\prime}) = \dsection(\dfunch(\dbasepoint^{\prime}))
\end{equation} 
where $\dfunch(\dbasepoint^{\prime}) = \dfunch(\dbasepoint)$.

\paragraph{\textbf{Records}}
\noindent
As introduced in \autoref{eq:artist:sheaves:monoid_morphism}, the fiber transformation $\dfunct: \dfuncpull \dtotal_{\dbasepoint^{\prime}} \rightarrow \dfuncpull \dtotal_{\dbasepoint^{\prime}}$ is a monoid action on the fiber $\dfunct \in Hom(\dfuncpull\dfiber\restriction_{\dbasepoint^{\prime}},\dfuncpull\dfiber\restriction_{\dbasepoint^{\prime}})$ restricted to a point $\dbasepoint^{\prime} \in \openset^{\prime}$. The fiber transformation is a change in section 
\begin{equation}
  \label{eq:atct:morphism:fiber}
  \dfunct: \dfuncpull \dsection \restriction_{\openset^{\prime}} \mapsto \dfuncpull \dsection^{\prime} \restriction_{\openset}
\end{equation}
where $\dsection, \dsection^{\prime} \in \Gamma(\openset^{\prime}, \dfuncpull\dtotal\restriction_{\openset^{\prime}})$. Since there are many different fiber transformations,  there is not a uniform way of expressing how this morphism acts on sections unlike in \autoref{eq:atct:morphism:verify_base}.\note{This is where we reintroduce the equivariance and homomorphism definitions from the introduction}

Examples of a fiber transformation are actions such as partial ordering for ranking purpose \cite{bruggemannRankingPrioritizationMultiindicator2011} and the permutation, ordering, translation, and scaling codified as Steven's measurement scales \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}.

@@

\begin{table}[h!]
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|lll|}\hline
      scale & property & scale specific constraints \\ \hline
      nominal & permutation &  $\dsectionc(\dbasepointc_1) \neq \dsectionc(\dbasepointc_2) \implies \dfunctc (\dsectionc(\dbasepointc_1)) \neq\dfunctc(\dsectionc(\dbasepointc_2)$)\\
      ordinal &  monotonic &  $\dsectionc(\dbasepointc_1) \leq \dsectionc(\dbasepointc_2) \implies \dfunctc (\dsectionc(\dbasepointc_1)) \leq \dfunctc(\dsectionc(\dbasepointc_2)$) \\
      interval &  translation &  $\dfunctc(\dsectionc(\dbasepointc) + C) = \dfunctc(\dsectionc(\dbasepointc)) + C$ \\
      ratio &  scaling &  $\dfunctc(\dsectionc(\dbasepointc) * C) = \dfunctc(\dsectionc(\dbasepointc))*C $\\ \hline
  \end{tabular}
  \caption{The record transformer $\dfunctc$ is defined by the general constraints listed in \autoref{tab:appendix:summary:datamod} and it must respect any properties that are specific to the record field. For example, this table lists the properties of the Steven's measurement scale types\cite{stevensTheoryScalesMeasurement1946}}
  \label{tab:appendix:summary:datatypes}
\end{table}

>>>>

The topological structure of the data can be expressed programmatically by constructing data types that encapsulate the class of topological structure-.i.e point, line, plane, network, etc. We propose that these data types can be formally specified as objects of a category. 
\begin{definition}\label{def:atct:category}
   An \textbf{category} $\mathcal{C}$ consists of the following \textit{data}: 
\begin{enumerate}
  \item a collection of \textit{objects} $X \textbf{ob}(\mathcal{C})$
  \item for every pair of objects $X, Y \in \textbf{ob}(\mathcal{C})$, a set of \textit{morphisms} $X \xrightarrow{f} Y \in Hom_{\mathcal{C}}(X, Y)$
  \item for every object $X$, a distinct \textit{identity morphism} $X \xrightarrow {id_x} X$ in $Hom_{\mathcal{C}}(X, X)$
  \item a \textit{composition function} $f \in Hom_{\mathcal{C}}(X, Y) \times  g \in Hom_{\mathcal{C}}(Y, Z) \rightarrow g \circ f \in Hom_{\mathcal{C}}(X, Z)$
\end{enumerate}
such that 
\begin{enumerate}
  \item \textit{unitality:} for every morphism $ X \xrightarrow{f} Y$, $f \circ id_x = f = id_y \circ f$
  \item \textit{associativity:} if any three morphisms $f, g, h$ are composable, 
    \begin{equation*}
      \begin{tikzcd}
        X \arrow[r, "f"] \arrow[rrr, "h\circ(g\circ f) = (h\circ g)\circ f"', bend right, dashed] & Y  \arrow[r, "g"] & Z \arrow[r, "h"] & W
        \end{tikzcd}
  \end{equation*}
  then they are associative such that $h\circ(g\circ f) = (h \circ g) \circ f$  \cite{lawvere2009conceptual,riehlCategoryTheoryContext,maclaneCategoriesWorkingMathematician2013,fongInvitationAppliedCategory2019}. 
  \end{enumerate}
\end{definition}
We encapsulate the topological structure of the data as a category $\mathcal{\dbasec}$. The standard construction of a category from a topological space is that it has open set objects $\opensetc$ and inclusion morphisms $\opensetc_i \xrightarrow{\iota} \opensetc_j$ such that $\opensetc_i \subseteq \opensetc_j$\cite{riehlCategoryTheoryContext}. The composability property expresses that inclusion is transitive, while associativity expresses that the inclusion functions can be curried in various equivalent groupings. By formally specifying the properties of the topological structure datatypes as $\mathcal{\dbase}$, we can express that these are the properties that are required as part of the implementation of the data type objects.

@@

\paragraph{\textbf{Topological structure}}
\noindent
The base space transformation is a pointwise continuous map from one open set to another open set in the same base space
\begin{equation}
  \label{eq:atct:morphism:base}
 \dfunchc: \dbasepointc^{\prime}\mapsto \dbasepointc
 \end{equation}
such that $\openset, \openset^{\prime} \subseteq \dbase$. This means $\openset$ and $\openset^{\prime}$ are of the same topology type. To correctly align the sections with the remapped base space, there is a a corresponding section pullback function
\begin{equation}
  \label{eq:atct:morphism:basepull}
  \dfuncpullc \dsectionc \restriction_{\opensetc^{\prime}}: \dsectionc\restriction_{\opensetc^{\prime}} \mapsto \dsectionc \restriction_{\opensetc^{\prime} \circ \dfunchc} 
\end{equation}
such that $\dsection|_{\openset} = \dfuncpull\dsection|_{\opensetc^{\prime}}$ because $\dsection|_{\openset} = \dsection|_{\dfunch(\opensetc^{\prime})}$. This means that the base space transformation  $\dfunch(\dbasepoint^{\prime}) = \dfunch(\dbasepoint)$ such that  
\begin{equation}
  \label{eq:atct:morphism:verify_base}
  \dsectionc(\dbasec) = \dfuncpullc\dsectionc(\dbasepointc^{\prime}) = \dsectionc(\dfunchc(\dbasepointc^{\prime}))
\end{equation} 
which means that the index of the record changes from $\dbasepoint$ to $\dbasepoint^{\prime}$ but the values in the record are unmodified.

\paragraph{\textbf{Records}}
As introduced in \autoref{eq:artist:sheaves:monoid_morphism}, the fiber transformation $\dfunctc$ is a change in section 
\begin{equation}
  \label{eq:atct:morphism:fiber}
  \dfunctc: \dfuncpullc \dsectionc \restriction_{\opensetc^{\prime}} \mapsto \dfuncpullc \dsectionc^{\prime} \restriction_{\opensetc}
\end{equation}
where $\dsection, \dsection^{\prime} \in \Gamma(\openset^{\prime}, \dfuncpull\dtotal\restriction_{\openset^{\prime}})$. Since $\dfunct$ maps from one continuous function to another, it must itself be continuous such that
\begin{equation}
  \label{eq:atct:morphism:fiber:continuity}
   \lim\limits_{x \rightarrow \dbasepointc^{\prime}}\dfunctc(\dfuncpullc\dsectionc(x)) =  \dfunctc(\dfuncpullc\dsectionc(\dbasepointc^{\prime}))
\end{equation}
As mentioned in \autoref{sec:atct:fb:fiber}, $\dfunct$ is also  
a morphism on the fiber category $\dfunct \in Hom(\dfuncpull\dfiber\restriction_{\dbasepoint^{\prime}},\dfuncpull\dfiber\restriction_{\dbasepoint^{\prime}})$ restricted to a point $\dbasepoint^{\prime} \in \openset^{\prime}$. This means $\dfunct$ has to statisfy the properties of a morphism (\autoref{def:atct:category})
\begin{itemize}
  \item \textit{closed}: $\dfunctc(\dfuncpullc\dsectionc(\dbasepointc^{\prime})) \in \dfiberc$
  \item \textit{unitality}: $\dfunctc(id_{\dfiberc}(\dfuncpullc\dsectionc(\dbasepointc^{\prime}))) = id_{\dfiberc}(\dfunctc(\dfuncpullc\dsectionc(\dbasepointc^{\prime})))$
  \item \textit{composition and associativity}: \\
  $\dfunctc(\dfunctc(\dfuncpullc\dsectionc(\dbasepointc^{\prime}))) = (\dfunctc\circ\dfunctc)(\dfuncpullc\dsectionc(\dbasepointc^{\prime}))$
\end{itemize}

Additionally, each fiber $\dfiber_{type}$ may have features in addition to the set of possible records. For example, monoids and groups are defined as sets with operators 
\note{clarify if $\dfunctc$ are actions on F or encode the operators/extra structure on F}

the monoids that underly partial ordering for ranking purpose \cite{bruggemannRankingPrioritizationMultiindicator2011} and the groups that define Steven's measurement scales \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}.

@@

\begin{table}[h!]
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|lll|}\hline
      scale & operators & sample constraint \\ \hline
      nominal & $=,\neq$ &  $\dsectionc(\dbasepointc_1) \neq \dsectionc(\dbasepointc_2)\implies \dfunctc (\dsectionc(\dbasepointc_1)) \neq\dfunctc(\dsectionc(\dbasepointc_2))$\\ 
      ordinal & $<, \leq, \geq, >$ &  $\dsectionc(\dbasepointc_1) \leq \dsectionc(\dbasepointc_2) \implies \dfunctc (\dsectionc(\dbasepointc_1)) \leq \dfunctc(\dsectionc(\dbasepointc_2)$) \\
      interval & $+, -$ &  $\dfunctc(\dsectionc(\dbasepointc) + C) = \dfunctc(\dsectionc(\dbasepointc)) + C$ \\
      ratio & $*,/$ &  $\dfunctc(\dsectionc(\dbasepointc) * C) = \dfunctc(\dsectionc(\dbasepointc))*C $\\ \hline
  \end{tabular}
  \caption{The record transformer $\dfunctc$ must satisfy the constraints listed in \autoref{tab:appendix:summary:datamod} and $\dfunctc$ must also respect the mathamatical structure of $\dfiber$. This table lists examples of $\dfunct$ preserving one of the binary operators that are part of the definition of each of the Steven's measurement scale types\cite{stevensTheoryScalesMeasurement1946}}. A full implementation would ensure that all operators that are defined as part of of $\dfiber$ are preserved.  
  \label{tab:appendix:summary:datatypes}
\end{table}
